Place of death in NRS COVID-19 deaths: data processing for analysis, 2022 data

### 2022-02-04 covid_nrs_place_of_death_data_2022.Rmd
nearly there with the data munging! can start visualising stuff soon I hope!

```{r}
weekly_cause_of_death_data <-
  bind_rows(
    weekly_cause_of_death_2021_onwards,
    weekly_deaths_cause_of_death_2020 %>% 
      select(-deaths_mean) %>%
      filter(cause_of_death != "All")  # don't need all cause data in causes breadkown!
  )

@@

>>>>

```{r}
weekly_deaths_cause_of_death <-
  bind_rows(
    weekly_cause_of_death_2021_onwards,
    weekly_deaths_cause_of_death_2020 %>% 
      select(-deaths_mean) %>%
      filter(cause_of_death != "All")  # don't need all cause data in causes breadkown!
  )

@@

# Merge weekly deaths with historic death rates

```{r}
## compute rolling average
merged_deaths_overall <-
  weekly_deaths_overall %>%
  filter(sex=="all" & age=="all") %>%
  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
  inner_join(historical_weekly_average_overall %>% mutate(sex = "all", age = "all"),  # only merge for aggregated data b/c that's all we have!
  by = c("place_of_death", "week_number", "sex", "age", "historic_period"))

merged_deaths_hb <-
  weekly_deaths_hb %>%
  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
  inner_join(historical_weekly_average_hb, by = c("week_number","ref_area","place_of_death", "historic_period"))

merged_deaths_la <-
  weekly_deaths_la %>%  # use data with weekly movingincluded
  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
  inner_join(historical_weekly_average_la, by = c("week_number","ref_area","place_of_death", "historic_period"))

merged_deaths_sex_age <-
  weekly_deaths_sex_age %>%  # use data with weekly movingincluded
  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
  inner_join(historical_weekly_average_sex_age, by = c("place_of_death", "week_number", "sex", "age", "historic_period"))

merged_deaths_sex <-
  weekly_deaths_sex %>%
  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
  inner_join(historical_weekly_average_sex, by = c("place_of_death", "week_number", "sex", "historic_period"))

merged_deaths_age <-
  weekly_deaths_age %>%
  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
  inner_join(historical_weekly_average_age, by = c("place_of_death", "week_number", "age", "historic_period"))

merged_deaths_cause_of_death <-
  weekly_deaths_cause_of_death %>%
  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
  inner_join(
    historical_average_cause_of_death, by = c("place_of_death", "week_number", "cause_of_death", "historic_period", "year", "date_w_c")
  )
```

## Weekly deaths including past weekly deaths

For this dataset, we simply `bind_rows` weekly deaths by year and place of death with past weekly deaths by year and location. Note that past data for years with 52 weeks includes an entry of 0 deaths for week 53, so we'll remove them, otherwise the date computation is thrown off!

```{r}
weekly_deaths_current_and_past <-
  bind_rows(
    weekly_deaths_overall %>%
    filter(sex=="all" & age=="all" & place_of_death!="All") %>%  # keep only aggregated values over sex & age
    select(year, week_number, week_number_run_over, place_of_death, matches("deaths"), -deaths_nrs_past_average_all_causes)
    ,
    past_deaths_overall %>%
      rename(deaths_all_causes = number_of_deaths) %>%
      filter(place_of_death!="All" & year!="average") %>%  # remove aggregated entries
      mutate(year = as.integer(year)) %>%
      filter(!(year %in% 2016:2019 & week_number==53))  # remove week 53 entries from years 2016-2019
  ) %>%
  mutate(date_w_c = compute_start_date_from_week_number(week_number = week_number, year_number = year))
```

# Annual deaths for 2020 & 2021

For visualising data using maps, we need to compute data for the entire year. Note: the non-covid and covid-related deaths in 2020 don't add up to all cause deaths due to the covid status only starting in week 12!

```{r}
# TODO: check through these to see if they compute correctly!

compute_total_deaths_in_year <- function(data_tbl, ...) {
  grouping_vars <- enquos(...)
  data_tbl %>%
    group_by(year, place_of_death, !!!grouping_vars) %>%
    summarise(across(.cols = c(deaths_all_causes, deaths_covid_related, deaths_non_covid), .fns = ~sum(.x,na.rm=TRUE)), .groups = "drop") %>%  # compute sums of all types of death
    mutate(deaths_all_causes_since_covid = deaths_covid_related + deaths_non_covid)  # compute a smaller number of all-cause deaths since covid started (over 2020, covid- and non-covid-related deaths don't add up to all-cause deaths because pre-covid deaths are included also!)
}

compute_ratio_of_annual_deaths_to_historical <- function(data_tbl) {
  data_tbl %>%
    mutate(
      ratio_annual_deaths_to_historical_mean = deaths_all_causes / annual_deaths_mean,
      ratio_annual_deaths_to_historical_min = deaths_all_causes / annual_deaths_min,
      ratio_annual_deaths_to_historical_max = deaths_all_causes / annual_deaths_max
    )
}

total_deaths_la <-
  merged_deaths_la %>%
  compute_total_deaths_in_year(ref_area) %>%
  left_join(historical_annual_average_la, by = c("place_of_death", "ref_area")) %>%
  compute_ratio_of_annual_deaths_to_historical

total_deaths_hb <-
  merged_deaths_hb %>%
  compute_total_deaths_in_year(ref_area) %>%
  left_join(historical_annual_average_hb, by = c("place_of_death", "ref_area")) %>%
  compute_ratio_of_annual_deaths_to_historical

total_deaths_sex <-
  merged_deaths_sex %>%
  compute_total_deaths_in_year(sex) %>%
  left_join(historical_annual_average_sex, by = c("place_of_death", "sex")) %>%
  compute_ratio_of_annual_deaths_to_historical

total_deaths_age <-
  merged_deaths_age %>%
  compute_total_deaths_in_year(age) %>%
  left_join(historical_annual_average_age, by = c("place_of_death", "age")) %>%
  compute_ratio_of_annual_deaths_to_historical

total_deaths_overall <-
  merged_deaths_overall %>%
  compute_total_deaths_in_year() %>%
  left_join(historical_annual_average_overall, by = c("place_of_death")) %>%
  compute_ratio_of_annual_deaths_to_historical

## bespoke computation because format is different... deaths split by cause!
total_deaths_cause_of_death <-
  merged_deaths_cause_of_death %>%
  group_by(year, place_of_death, cause_of_death) %>%
  summarise(deaths = sum(number_of_deaths), .groups="drop") %>%
  left_join(historical_annual_average_cause_of_death, by = c("place_of_death", "cause_of_death")) %>%
  mutate(
    ratio_annual_deaths_to_historical_mean = deaths / annual_deaths_mean
  )

current_and_historical_deaths_by_cause_and_location <-
  bind_rows(
    ## 2020 & 2021 separately
    total_deaths_cause_of_death %>%
      select(year, place_of_death, cause_of_death, deaths) %>%
      mutate(year = as.character(year)),
    ## 2020 & 2021 averaged
    total_deaths_cause_of_death %>%
      select(year, place_of_death, cause_of_death, deaths) %>%
      group_by(place_of_death,  cause_of_death) %>%
      summarise(deaths = sum(deaths), year = "2020-2021", .groups = "drop"),
    ## historical data
    total_deaths_cause_of_death %>%
      select(year, place_of_death, cause_of_death, deaths = annual_deaths_mean) %>%
      mutate(year = "2015-2019") %>%
      distinct
  ) %>%
  filter(place_of_death != "All") %>%
  group_by(year, cause_of_death) %>%
  mutate(proportion = deaths / sum(deaths)) %>%
  ungroup %>%
  arrange(year, cause_of_death)

```
